Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2025-07-09 _manuscript/revision/response.qmd

---
title: "Response to Reviewers JDS24112"
subtitle: "Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks"
author: "H. Sherry Zhang, Roger D. Peng"
date: "2025-07-08"
format: 
  html:
    toc: false
---

We would like to thank the reviewers and editor for taking the time to review our paper and provide constructive comments. Below is a point-by-point explanation of how we have changed the paper. Reviewers' comments are in normal text and **our responses are in bold text**.

# Reviewer 1

## General comments

The manuscript is being considered for the special issue on Statistical Aspects of Trustworthy Machine Learning. However, the connection to the theme of the special issue feels somewhat underdeveloped. It would be helpful for the authors to clarify which aspect(s) of trustworthiness are being addressed. While the work appears to relate to robustness and reproducibility, these connections should be more clearly articulated and substantiated in the manuscript.

The proposed procedure appears to assume that it is possible to simulate datasets similar to the observed data (e.g., as illustrated in Figure 2). However, this assumption may not be realistic in many practical scenarios. Simulating data requires a data-generating model, but commonly used statistical methods such as generalized estimating equations (GEE) or the generalized method of moments (GMM) do not fully specify the underlying distribution or likelihood. Even when the mean structure is specified, there may be infinitely many likelihood models consistent with it. Simulating from only one such model risks overlooking a broad class of plausible alternatives. How do the authors propose to handle such cases? More broadly, generating synthetic data with properties similar to real data is a fundamental and challenging problem in data science practice. The authors may wish to comment on recent advances in this area that aim to address this challenge without fully specifying a probabilistic model.

The construction of validation checks appears to be a critical step in applying the proposed methodology. It would be valuable if the authors could offer some guidance or recommendations on the principles for designing these checks. Clarifying this aspect would enhance the practical utility of the approach and help readers apply it more effectively in real-world settings.

The independence metric appears to be ill-defined or uninformative for single checks, much like defining total correlation for a single variable. By definition, the metric seems to always take the value one in such cases, as observed in Tables 1 and 3. Including it in the total score may artificially inflate the overall score when compared to checks involving multiple components. Further, in the current examples, the multivariate checks consistently yield independence values close to one. Are there cases where this metric meaningfully deviates from one? Clarifying the interpretation and role of this metric, particularly in the single-check case, would strengthen the methodology.

The writing is generally clear and easy to follow. However, there are a number of minor syntax errors that could be addressed with more careful proofreading. The use of AI-based writing assistance tools could be helpful in this regard and, in my view, would constitute an ethical and constructive application of such tools. **A: We have run though the spelling check on the article and fixed the spelling mistakes.**

## Specific Comments

1.  Section 4, paragraph 2: Should the “simulated datasets” be independent of the observed data? If so, this is a big difference from bootstrap samples and a remark would be useful.

2.  Figure 2: Quadratic mean is mentioned in the text but not shown in the SCORE box. **A: easy.**

3.  Section 4.1, paragraph 1: X1, . . . , Xn’s are undefined at the first occurrence; n means the number of checks, but in statistics it is often used for sample size. **A: easy. TODO**

4.  Section 4.2, end of paragraph 1: I assume the dash should be the minus sign. Having a minus sign in a sentence is confusing. This needs to be rephrased, perhaps just use “minus”.  **A: We have replaced the dash with the word "minus" in the text. **

5.  Section 4.2, paragraph 2: What is the range of total correlation? My understanding is that it is \[0, ∞). Even after standardizing to per observation, the range may still be not \[0, 1) as desired. **A: The range of total correlation is \[0, ∞), to obtain a measure in \[0, 1), we calculate total correlation per observation as the independence score. This is explained in the last sentence of the paragraph. **

6.  Continue: p is undefined; the equation needs to close with a period.  **A: We have added the definition of p(.) as TODO and close the sentence.**

7.  Table 2: How did the authors come up with these checks? Readers would appreciate the motivations of these checks. Some of the checks are apparently correlated (e.g., the four checks on mortality-temperature correlation).

8.  Section 5.1, paragraph 1: Gaussian copula does not allow tail dependence. So by imposing a Gaussian copula, it is assumed that there is no tail dependence in these three variables. As questioned in one of my general comments, this might not be desirable.

9.  Figure 4: Could add bands of 95% confidence intervals of the QQ-plot; the texts on top of each panel could be moved to the caption; the labels of the axis should clarify empirical versus theoretical.

10. Figure 5 caption: The negative sign in front of 0.03 needs to be fixed. **A: Fixed.**

11. Section 5, paragraph 1: The “our second example” appears to be a continuation of the same analysis. **A: The intention of using "second example" is because we count the toy example as the first example. But to avoid this confusion, we have changed this to "In the following example".**

12. Table 4: There are four checks for the mortality-temperature correlation, but only one check reported in this table.

13. Section 5: It’d be valuable to see some real improvements from the checks. For example, fit a poor model first and identify the issues from the checks; then fit a strong model to fix the issues and pass the checks.

## Minor comments

-   Abstract: “simulations of the original data” is misleading. **A: we remove this information in this sentence for easier understanding of the sentence. **

-   Page 2, paragraph 1: “thought” process, not “throught” process. **A: Fixed.**

-   Page 2, paragraph 2: “is measures”, two verbs. **A: The redundant "is" has been removed**.

### 2025-07-09 _manuscript/_tex/index.tex
some revision

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{interact}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks},
  pdfauthor={H. Sherry Zhang; Roger D. Peng},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

We might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as, TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the analysts.
Ideally, there could be a way to explicitly communciate these throught
processes in the data analysis to others. Even better, if these
expectations were machine-readable, we could analyze them and learn from
the analysis itself. For example, we could answer questions about
whether the checks also apply to other researchers analyzing new data in
the same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand. Externalizing
these thought processes is a practice that has potential to improve the
trustworthiness of the analyses \citep{yu2024veridical} in general and
the trustworthiness of subsequent products, such as machine learning
models, that may be built on such analyses. Developing an approach to
reveal some of this process without requiring a reader to essentially
reconstruct the analysis process from the raw data would provide an
improved basis for trusting an analysis result
\citep{peng2021reproducible}.

In this paper, we conceptualize these inexplicit expectations as
\emph{analysis validation checks}, which allows us to examine the
assumptions made about the data during an analysis. We then introduce a
procedure to identify the subset of checks that best predict the
occurrence of unexpected outcomes. The procedure, based on simulations
of the original data, compares the accuracy and independence of
different combinations of analysis validation checks. Accuracy is
determined using binary classification metrics (precision and recall)
from a logic regression model \citep{ruczinski_logic_2003}, while
independence is measures the shared information among checks. The
proposed workflow offers a numerical guarantee that the analysis will
produce the expected results, assuming the assumptions about the data
generating mechanism hold.

@@

The concept of data quality has been studied in the literature, with
earlier work focusing on defining frameworks -- such as dimensions,
attributes, and measures -- to improve data quality in database and
information systems
\citep{wang1996beyond, batini2009methodologies, 6204995, woodall2014classification, cai2015challenges, 8642813}.
More recently, data validation has been incorporated into frameworks
like Google TensorFlow \citep{polyzotis2019data} to ensure the quality
of data for training machine learning models, as well as for supporting
business decision-making \citep{schelter2018automating}. With the
increasing availability of open data in scientific research, the users
of the data are often not the original data collector and may not be
aware of all the details or nuances of the data. This encourages
researchers to conduct data quality checks before beginning the analytic
process, helping to avoid unexpected discoveries later on. In R,
packages like \texttt{skimr} \citep{skimr} and \texttt{dataMaid}
\citep{dataMaid} provide basic data screening and data quality reports,
while packages such as \texttt{assertr} \citep{assertr},
\texttt{validate} \citep{validate}, and \texttt{pointblank}
\citep{pointblank} focuse on providing infustructures that allow users
to define customized data quality checks.

@@

While not all analysis validation checks are equally important, some may
be more useful at diagnosing unexpected outcomes than others. This
section introduces a procedure to identify a subset of analysis
validation checks that best predict the occurence of unexpected
outcomes.

@@

From the simulated data, the accuracy branch in
Figure~\ref{fig-metric-calc} refers to the set of checks' ability to
detect unexpected outcomes accurately. To relate multiple checks to the
outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
Originally developed for SNP microarray data, logic regression
constructs Boolean combinations of binary variables,
\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict classification or
regression outcomes. The logical operations allowed to construct
\(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
(\(\neg\)) and an example of the Boolean combination of
\(\mathcal{L(\cdot)}\) is \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\). In
our binary-binary classification problem, the logic regression model can
be written as follows:

@@

This objective function is optimized using simulated annealing algorithm
to minimize the misclassification error and the following 6 moves are
permitted during the optimization to grow or purne the tree (as detailed
in Figure 2 of \citet{ruczinski_logic_2003}): 1) replacing a leaf node,
2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5)
splitting a leaf node, and 6) deleting a leaf node. Predictions from the
logic regression model is then used, along with the observed outcome, to
calculate the precision and recall of the optimized Boolean combination
of checks. Compared to other tree-based methods for binary-binary
prediction, the Boolean combinations from the logic regression model
produce a tree structure that can be directly interpreted as the
possible combination of checks leading to an unexpected outcome, without
the need to invert the tree as required in classic tree-based recursive
partitioning methods.

@@

While checks may score high on predictive accuracy, they may be
correlated and less useful to provide actions to act on the analysis.
This could happen if a set of checks are all tangentially related to the
cause of the unexpected results, but none addresses the root cause. The
independence score in Figure~\ref{fig-metric-calc} is defined as 1 -
total correlation per observation.

@@

\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)}\]

A high total correlation value indicates overlapping of information
among the analysis validation checks, while a low value suggests that
the checks are independent and provides unique information to diagnose
the unexpected outcome. To standardize this measure, the total
correlation \emph{per observation} is used in calculating the
independence score.

The three metrics (precision, recall, and independence) can be combiend
into a single metric using the arithmetic mean, harmonic mean, or
quadratic mean. The differences among these means are minimal when the
three metrics are similar. However, as the differences among the metrics
increases, the harmonic mean tends to produce the smallest overall
score, as it penalizes low values, while the quadratic mean tends to
produce the largest score by rewarding higher values more. For simple
interpretation of the score, the arithmetic mean is preferred, while in
applications where the difference between precision, recall, and
independence need to be penalized or rewarded more, the harmonic and
quadratic mean should be considered.

@@

In the study of the health effects of outdoor air pollution, one area of
interest is the association between short-term, day-to-day changes in
particulate matter air pollution and daily mortality counts. Substantial
work has been done to study this question and to date, there appears to
be strong evidence of an association between particulate matter less
than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
mortality from all non-accidental causes \citep{samet2000fine}. For our
second example, we use the problem of studying PM10 and mortality along
with data from the National Morbidity, Mortality, and Air Pollution
Study (NMMAPS) to demonstrate how our analysis validation checks
described in Section~\ref{sec-method} can be applied. In addition to
providing a more substantial problem for our methods, this example also
demonstrates how the procedure presented in Section~\ref{sec-method} can
be used to select cutoff values in the analysis checks to diagnose an
unexpected PM10 coefficient from the generalized linear model. The
dataset that we use to develop our simulations and analysis checks is
from New York City, and contains daily PM10, all-cause (non-accidental)
mortality, and average temperature values from 1992--2000. We then will
apply the analysis checks to similar data from 39 other cities from the
NMMAPS study.

@@

\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/unnamed-chunk-8-1.pdf}}

@@

\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
fourteen checks and the outcome expectation (unexpected) as the response
variable. The model suggests the relationship: (mortality-PM10
correlation \textless{} -0.03) AND (mortality-temperature correlation
\textgreater{} -0.3 OR (there exist mortality outlier AND there exist
PM10 outlier)) to predict the unexpected PM10 coefficient.}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{interact}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[]{natbib}
\bibliographystyle{plainnat}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks},
  pdfauthor={H. Sherry Zhang; Roger D. Peng},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

We might gain insight into analysts' thought processes by speaking with
them directly or watching them work via screencast videos they produce,
such as, TidyTuesday screencast videos or think-aloud type studies
\citep[e.g.][]{gu2024data}. However, direct observation of analysis is
not scalable and may not always be feasible; creating educational
screencast videos requires significant effort from the analysts.
Ideally, there could be a way to explicitly communicate these thought
processes in the data analysis to others. Even better, if these
expectations were machine-readable, we could analyze them and learn from
the analysis itself. For example, we could answer questions about
whether the checks also apply to other researchers analyzing new data in
the same context, whether they reflect common practices in the field, or
whether they are specific to the data or analysis at hand. Externalizing
these thought processes is a practice that has potential to improve the
trustworthiness of the analyses \citep{yu2024veridical} in general and
the trustworthiness of subsequent products, such as machine learning
models, that may be built on such analyses. Developing an approach to
reveal some of this process without requiring a reader to essentially
reconstruct the analysis process from the raw data would provide an
improved basis for trusting an analysis result
\citep{peng2021reproducible}.

In this paper, we conceptualize these implicit expectations as
\emph{analysis validation checks}, which allows us to examine the
assumptions made about the data during an analysis. We then introduce a
procedure to identify the subset of checks that best predict the
occurrence of unexpected outcomes. The procedure, based on simulations
of the original data, compares the accuracy and independence of
different combinations of analysis validation checks. Accuracy is
determined using binary classification metrics (precision and recall)
from a logic regression model \citep{ruczinski_logic_2003}, while
independence measures the shared information among checks. The proposed
workflow offers a numerical guarantee that the analysis will produce the
expected results, assuming the assumptions about the data generating
mechanism hold.

@@

The concept of data quality has been studied in the literature, with
earlier work focusing on defining frameworks -- such as dimensions,
attributes, and measures -- to improve data quality in database and
information systems
\citep{wang1996beyond, batini2009methodologies, 6204995, woodall2014classification, cai2015challenges, 8642813}.
More recently, data validation has been incorporated into frameworks
like Google TensorFlow \citep{polyzotis2019data} to ensure the quality
of data for training machine learning models, as well as for supporting
business decision-making \citep{schelter2018automating}. With the
increasing availability of open data in scientific research, the users
of the data are often not the original data collector and may not be
aware of all the details or nuances of the data. This encourages
researchers to conduct data quality checks before beginning the analytic
process, helping to avoid unexpected discoveries later on. In R,
packages like \texttt{skimr} \citep{skimr} and \texttt{dataMaid}
\citep{dataMaid} provide basic data screening and data quality reports,
while packages such as \texttt{assertr} \citep{assertr},
\texttt{validate} \citep{validate}, and \texttt{pointblank}
\citep{pointblank} focus on providing infrastructures that allow users
to define customized data quality checks.

@@

While not all analysis validation checks are equally important, some may
be more useful at diagnosing unexpected outcomes than others. This
section introduces a procedure to identify a subset of analysis
validation checks that best predict the occurrence of unexpected
outcomes.

@@

From the simulated data, the accuracy branch in
Figure~\ref{fig-metric-calc} refers to the set of checks' ability to
detect unexpected outcomes accurately. To relate multiple checks to the
outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
Originally developed for SNP micro-array data, logic regression
constructs Boolean combinations of binary variables,
\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict classification or
regression outcomes. The logical operations allowed to construct
\(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
(\(\neg\)) and an example of the Boolean combination of
\(\mathcal{L(\cdot)}\) is \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\). In
our binary-binary classification problem, the logic regression model can
be written as follows:

@@

This objective function is optimized using simulated annealing algorithm
to minimize the misclassification error and the following 6 moves are
permitted during the optimization to grow or prune the tree (as detailed
in Figure 2 of \citet{ruczinski_logic_2003}): 1) replacing a leaf node,
2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5)
splitting a leaf node, and 6) deleting a leaf node. Predictions from the
logic regression model is then used, along with the observed outcome, to
calculate the precision and recall of the optimized Boolean combination
of checks. Compared to other tree-based methods for binary-binary
prediction, the Boolean combinations from the logic regression model
produce a tree structure that can be directly interpreted as the
possible combination of checks leading to an unexpected outcome, without
the need to invert the tree as required in classic tree-based recursive
partitioning methods.

@@

While checks may score high on predictive accuracy, they may be
correlated and less useful to provide actions to act on the analysis.
This could happen if a set of checks are all tangentially related to the
cause of the unexpected results, but none addresses the root cause. The
independence score in Figure~\ref{fig-metric-calc} is defined as 1 minus
total correlation per observation.

@@

\[C(X_1, X_2, \cdots, X_n) = \sum_{x_1} \sum_{x_2} \cdots \sum_{x_n} p(x_1, x_2, \cdots, x_n) \log \frac{p(x_1, x_2, \cdots, x_n)}{p(x_1)p(x_2) \cdots p(x_n)},\]

where \(p(.)\) is the TODO. A high total correlation value indicates
overlapping of information among the analysis validation checks, while a
low value suggests that the checks are independent and provides unique
information to diagnose the unexpected outcome. To standardize this
measure, the total correlation \emph{per observation} is used in
calculating the independence score.

The three metrics (precision, recall, and independence) can be combined
into a single metric using the arithmetic mean, harmonic mean, or
quadratic mean. The differences among these means are minimal when the
three metrics are similar. However, as the differences among the metrics
increases, the harmonic mean tends to produce the smallest overall
score, as it penalizes low values, while the quadratic mean tends to
produce the largest score by rewarding higher values more. For simple
interpretation of the score, the arithmetic mean is preferred, while in
applications where the difference between precision, recall, and
independence need to be penalized or rewarded more, the harmonic and
quadratic mean should be considered.

@@

In the study of the health effects of outdoor air pollution, one area of
interest is the association between short-term, day-to-day changes in
particulate matter air pollution and daily mortality counts. Substantial
work has been done to study this question and to date, there appears to
be strong evidence of an association between particulate matter less
than 10 \(\mu\)g/m\(^3\) in aerodynamic diameter (PM10) and daily
mortality from all non-accidental causes \citep{samet2000fine}. In the
following example, we use the problem of studying PM10 and mortality
along with data from the National Morbidity, Mortality, and Air
Pollution Study (NMMAPS) to demonstrate how our analysis validation
checks described in Section~\ref{sec-method} can be applied. In addition
to providing a more substantial problem for our methods, this example
also demonstrates how the procedure presented in
Section~\ref{sec-method} can be used to select cutoff values in the
analysis checks to diagnose an unexpected PM10 coefficient from the
generalized linear model. The dataset that we use to develop our
simulations and analysis checks is from New York City, and contains
daily PM10, all-cause (non-accidental) mortality, and average
temperature values from 1992--2000. We then will apply the analysis
checks to similar data from 39 other cities from the NMMAPS study.

@@

\pandocbounded{\includegraphics[keepaspectratio]{index_files/figure-pdf/unnamed-chunk-3-1.pdf}}

@@

\caption{\label{fig-linear-reg-tree}Logic regression model fitted to the
fourteen checks and the outcome expectation (unexpected) as the response
variable. The model suggests the relationship: (mortality-PM10
correlation \textless{} \(-0.03\)) AND (mortality-temperature
correlation \textgreater{} \(-0.3\) OR (there exist mortality outlier
AND there exist PM10 outlier)) to predict the unexpected PM10
coefficient.}
